Splitting it up: the \pkg{spduration} split-population duration regression package

### 2015-11-03 manuscript.Rmd
spell checking & bibs

---
author:
  - name: Daniel W. Hill
    affiliation: University of Georgia
    address: >
      First line
      Second line
    email: dwhill@uga.edu
    url: http://spia.uga.edu/directory/faculty_staff/hill-daniel
  - name: Nils W. Metternich
    affiliation: University College London
  - name: Andreas Beger
    affiliation: Duke University
  - name: Michael D. Ward
    affiliation: Duke University
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration model. The package includes estimation and forecasting methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{R}"]
  plain:     [keywords, not capitalized, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

This template demonstrates some of the basic latex you'll need to know to create a JSS article.

## Code formatting

Don't use markdown, instead use the more precise latex commands:

* \proglang{Java}
* \pkg{plyr}
* \code{print("abc")}

@@

This insight was realized as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients following treatment, and where quite obviously some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds. As with conventional duration models, their origin in health and medicine has shaped the terminology conventionally used with such model, e.g. split-population duration models are sometimes also referred to in such contexts as cure rate models, and the basic concepts like survival and failure rates reference the survival of humans.

Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine. In an early and foundational application that reached beyond criminal science, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints. This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.

In public health, @douglas1994hazard use data from the US to model factors related to the age at which smokers started the habit, and @forster2001role examines the impact of tobacco taxes on smoking and quitting decisions. @deyoung2003failure models the failure of new commercial banks in the US during the 1980's. Lastly, in our own domain, political science, @svolik2008authoritarian used a split-population duration framework to study democratic consolidation, i.e. whether democratic regimes persist or slide back to authoritarianism, and when. Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project [@ward2013learning] and to model irregular leadership changes for the Political Instability Task Force [PITF; @beger2014ensemble].

@@

The last component to complete the likelihood is the choice of a distribution for the shape of the hazard rate. The \pkg{spduration} package implements two hazard rate shapes, Weibull\footnote{In the code, $f(t) = \alpha \lambda^{\alpha} t^{\alpha - 1} e^{-(\lambda t)^\alpha}$.} and log-logistic:

@@

Using the \code{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values. These can be exported by calling \code{xtable} directly on the "spdur" object to produce Table \ref{loglog-table}.

@@

One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects.

@@

The table shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation. Interestingly, the coup risk variable has a rather large and statistically significant coefficient in the risk equation.

@@

```{r hazard-ex-1, eval = FALSE}
plot(loglog_model, type="hazard", main = "A")
```

@@

```{r hazard-ex-2, eval = FALSE}
plot(loglog_model, type="hazard", 
     xvals = c(1, 1, 10, 0.05), 
     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
     main = "B")
```

Importantly, the package also includes functions to evaluate model predictions. The generic function \code{predict} can be used on an object of class \code{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period. In addition, the generic \code{plot} function can be used on \code{spdur} objects to produce a separation plot [@greenhill2011separation], which is a graphical display for evaluating model predictions. The code below produces Figure \ref{insamp}.

@@

@@

Lastly, here is a comparision to regular logistic regression.

```{r logistic regression}
glm_model2    <- glm(
  coup ~ milreg + instab + regconf + couprisk + wealth + rwar + regconf + 
    samerica + camerica,
  data = coup_train, family = "binomial"
)

c(Weibull = AIC(weib_model2), 
  Loglog = AIC(loglog_model2), 
  Logistic = AIC(glm_model2))
```

```{r, eval = FALSE, include = FALSE}
glm2_test_p   <- predict(glm_model2, newdata = coup_test, type = "response")
glm2_test_p   <- glm2_test_p[!is.na(glm2_test_p)]

obs_y <- coup_test[complete.cases(coup_test), "coup"]

library("separationplot")
separationplot(glm2_test_p,   obs_y)

brier(obs_y, glm2_test_p)
auc_roc(obs_y, glm2_test_p)
auc_roc(obs_y, weib2_test_p)
auc_roc(obs_y, loglog2_test_p)
auc_pr(obs_y, glm2_test_p)
auc_pr(obs_y, weib2_test_p)
auc_pr(obs_y, loglog2_test_p)
```

>>>>

---
author:
  - name: Daniel W. Hill
    affiliation: University of Georgia
    address: >
      First line
      Second line
    email: dwhill@uga.edu
    url: http://spia.uga.edu/directory/faculty_staff/hill-daniel
  - name: Nils W. Metternich
    affiliation: University College London
  - name: Andreas Beger
    affiliation: Duke University
  - name: Michael D. Ward
    affiliation: Duke University
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration models. The package includes estimation and several post-estimation methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [regression, rare events, "\\proglang{R}"]
  plain:     [regression, rare events, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

This insight was realized as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients following treatment, and were quite obviously some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds. As with conventional duration models, their origin in health and medicine has shaped the terminology conventionally used with such model, e.g. split-population duration models are sometimes also referred to in such contexts as cure rate models, and the basic concepts like survival and failure rates reference the survival of humans.

Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine. In an early and foundational application that reached beyond medicine, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal lapses at all, and if so which factors are related to the amount of time between prison stints. This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.

In a complementary set of research from public health, @douglas1994hazard use data from the US to model the age at which smokers started the habit, and @forster2001role examine the impact of tobacco taxes on smoking and quitting decisions. @deyoung2003failure, an economist, models the failure of new commercial banks in the US during the 1980's. Lastly, in our own domain, political science, @svolik2008authoritarian used a split-population duration framework to whether democratic regimes persist or slide back to authoritarianism, and when. Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project [@ward2013learning] and to model irregular leadership changes for the Political Instability Task Force [PITF; @beger2014ensemble].

@@

The last component to complete the likelihood is the choice of a distribution for the shape of the hazard rate. The \pkg{spduration} package implements two hazard rate shapes, Weibull and log-logistic:

@@

Using the \code{summary} function on either model object will produce standard output showing the model formula, estimates for the duration and risk equations, and test statistics with $p$-values. These can be exported by calling \code{xtable} directly on the "spdur" object to produce Table 2.

@@

The table shows estimates from the duration equation, beginning with the intercept, and then estimates from the risk equation. The duration component of the model is in accelerated failure time form and the effect of coefficient estimates is on the log of expected time to failure. The negative coefficient for military regimes for example shows that the expected time to a coup is shorter in military regimes than non-military regimes, holding all other factors constant. In the risk equation positive coefficients mean higher risk of coup, and military regimes thus also have a higher risk of experiencing any coup at all.

One may also use the \code{AIC} and \code{BIC} function to calculate the information criterion statistics for \code{spdur} objects. In our example both models are close enough in both statistics to be indistinguishable, so we will continue to focus on the log-logistic form.

@@

@@

```{r hazard-ex-1, eval = FALSE}
plot_hazard(loglog_model, main = "A")
```

@@

As mean values are maybe not that interesting, we can also supply our values for a scenario of choice. The values are used in the same order that variables are specified in the equations used to estimate the models, which also corresponds to the order of variables in Table 2. Including an intercept term. This produces plot B in Figure \ref{hazard-ex}

```{r hazard-ex-2, eval = FALSE}
plot_hazard(loglog_model, 
     xvals = c(1, 1, 10, 0.05), 
     zvals = c(1, 7, 8.64, 1, 1, 0.05, 0, 0),
     main = "B")
```

The hazard plots can also be called with \code{plot(..., type = "hazard")}. Without a type argument, \code{plot.spdur} will produce a separation plot, which is a graphical display for evaluating model predictions [@greenhill2011separation]. The code below produces Figure \ref{insamp}.

@@

Underlying both plotting functions is the \code{predict} function, which can be used on an object of class \code{spdur} to generate several kinds of predictions, including the probability that an observation is ``at-risk'' and the probability of failure for a given time period.

@@
